Linear analysis of the vertical shear instability in astrophysical disks

### 2014-12-29 linear_adia.tex
new fig for analytic discussion with therm relax

Note that when $\gamma>1$ and $q=0$, the term with $\ln\rho^{\prime
  2}$ in the integrand (third term on the RHS) represents
stabilization by the vertical entropy gradient (real bouyancy frequency). 
It gives an imaginary contribution to $\hat{\sigma}$ when $q\neq0$,
but this is a factor $|\epsilon/\hat{k}|$ smaller than the real
(stabilizing) contribution, 
since $\epsilon\ll 1$ for a thin disk and we (usually) consider
$\hat{k}\gg1$. Thus, instability is expected to be associated with the 
last term on the RHS of Eq. \ref{adia_integral}.

@@

\subsubsection{The case of $\gamma=2$}
In principle, we can solve Eq. \ref{adia_disp} for the mode
frequencies and growth rates as a function of the disk parameters and
the radial wavenumber. This is complicated in 
general, but simplifies for $\gamma=2$, for which the quantity
$A^2+4C$ is real. In this case we find
\begin{align}
  \hat{\nu}^2 =
  \frac{1}{2}\frac{(2M+1)}{\sqrt{1+2\hat{k}^2}}&\left[
    \sqrt{1-\frac{4M(M+1)\hat{k}^2\epsilon^2q^2}{(2M+1)^2(1+2\hat{k}^2)}}\right.\notag\\
  &\phantom{=}
  \left.- \sqrt{1 - \frac{\hat{k}^2\epsilon^2q^2}{(1+2\hat{k}^2)}}\right].
\end{align}
In the limit of a thin disk $\epsilon\to0$ or weak shear $|q|\to0$, we
have 
\begin{align}
  \hat{\nu}^2 = \frac{\epsilon^2
    q^2}{4(2M+1)}\frac{\hat{k}^2}{(1+2\hat{k}^2)^{3/2}},\label{gam2_growth}
\end{align}
and the maximum growth rate occurs at $\hat{k}=\hat{k}_\mathrm{opt}=1$,  
\begin{align}
  \mathrm{max}(\hat{\nu}) &= \frac{|\epsilon
    q|}{2(3^{3/4}\sqrt{2M+1})}\label{gam2_max_growth}.
  % & \leq \frac{|\epsilon
  % q|}{2\times3^{3/4}}\notag.
\end{align}
Unlike isothermal perturbations, Eq. \ref{gam2_max_growth} shows that
there is a maximum possible growth rate for $M=0$, and growth rates decrease with increasing $M$. This
is despite the fact that in the thin-disk approximation 
$d\Omega^2/dz\propto z$, so there is infinite vertical shear
available.

@@

Reversing the argument, we expect to recover the VSI when the bouyancy
frequency becomes negligible compared to the cooling rate, or 
\begin{align}
  t_c \ll \frac{\Omega}{\gamma N_z^2}. 
\end{align} 
Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
for $N_z^2$ appropriate for vertically isothermal disks, we expect
\begin{align}
  \beta \ll
  \frac{1}{H^2_\mathrm{iso}\left(\gamma-1\right)\left(\frac{\p\ln\rho}{\p
          z}\right)^2} 
\end{align}
is required to enable the VSI, where we used $\Omega\sim\Omega_k$.  
This inequality is satisfied at the
midplane and as $|z|\to\infty$. It can be satisifed at all heights if
it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
Using the full expression for the density field for vertically isothermal
disks (Eq. \ref{vertiso_eqm}), we expect 
\begin{align}
  \beta \ll \frac{27}{4(\gamma-1)}\epsilon^2,
\end{align}
i.e. $\beta\ll O(\epsilon^2)$, is a sufficient condition for
bouyancy to be negligible.

@@

One effect of introducing finite thermal relaxation is buoyancy. % We
% show in the next section that buoyancy strongly stabilizes the
% VSI.
The $\zhat^2$ depedence in $\delta v_{z1}$ then makes sense because it
is most significant for large $\zhat$, i.e. away from the midplane
where the effect of bouyancy first appears as $t_c$ is increased.

@@

It is simplest to solve Eq. \ref{relax_disp} numerically. In
Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
($M=0$) VSI mode as a function of $k_xH_\mathrm{iso}$ for a range of
dimensionless thermal relaxation timescales $\beta$, in a disk with
$\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. Consistent with the previous
discussion, for $\beta\to0$ thermal relaxation has no effect as
$k_x\to0$ or $k_x\to\infty$.

Fig. \ref{relax_disp_fig} shows that a growing fundamental VSI mode
exists both in the isothermal ($\beta\to0$) and adiabatic
($\beta\to\infty$) limits, although the latter growth rates are
significantly smaller and perturbations are restricted to the
midplane. However, for $0.1 \lesssim \beta\lesssim
\beta 30$, this mode is stabilized for $k_xH_\mathrm{iso}\gtrsim
10$. Although results for small $k_x$ violates the low-frequency
approximation, Fig. \ref{relax_disp_fig} suggests that thermal
relaxation needs to be significantly shorter than the dynamical time
in order for the VSI with appreciable growth rates to develop.

@@

>>>>

Note that when $\gamma>1$ and $q=0$, the term with $\ln\rho^{\prime
  2}$ in the integrand (third term on the RHS) represents
stabilization by the vertical entropy gradient (real buoyancy frequency). 
It gives an imaginary contribution to $\hat{\sigma}$ when $q\neq0$,
but this is a factor $|\epsilon/\hat{k}|$ smaller than the real
(stabilizing) contribution, 
since $\epsilon\ll 1$ for a thin disk and we (usually) consider
$\hat{k}\gg1$. Thus, instability is expected to be associated with the 
last term on the RHS of Eq. \ref{adia_integral}.

@@

\subsubsection{The case of $\gamma=2$}
In principle, we can solve Eq. \ref{adia_disp} for the mode
frequencies and growth rates as a function of the disk parameters and
the radial wavenumber. This is complicated in 
general, but simplifies for $\gamma=2$, for which the quantity
$A^2+4C$ is real. In this case we find
\begin{align}
  \hat{\nu}^2 =
  \frac{1}{2}\frac{(2M+1)}{\sqrt{1+2\hat{k}^2}}&\left[
    \sqrt{1-\frac{4M(M+1)\hat{k}^2\epsilon^2q^2}{(2M+1)^2(1+2\hat{k}^2)}}\right.\notag\\
  &\phantom{=}
  \left.- \sqrt{1 - \frac{\hat{k}^2\epsilon^2q^2}{(1+2\hat{k}^2)}}\right].
\end{align}
In the limit of a thin disk $\epsilon\to0$ or weak shear $|q|\to0$, we
have 
\begin{align}
  \hat{\nu}^2 = \frac{\epsilon^2
    q^2}{4(2M+1)}\frac{\hat{k}^2}{(1+2\hat{k}^2)^{3/2}},\label{gam2_growth}
\end{align}
and the maximum growth rate occurs at $\hat{k}=\hat{k}_\mathrm{opt}=1$,  
\begin{align}
  \mathrm{max}(\hat{\nu}) &= \frac{|\epsilon
    q|}{2(3^{3/4}\sqrt{2M+1})}\label{gam2_max_growth}.
  % & \leq \frac{|\epsilon
  % q|}{2\times3^{3/4}}\notag.
\end{align}
Unlike isothermal perturbations, Eq. \ref{gam2_max_growth} shows that
maximum growth rate occurs at $M=0$, and growth rates decrease with increasing $M$. This
is despite the fact that in the thin-disk approximation 
$d\Omega^2/dz\propto z$, so there is infinite vertical shear
available.

@@

Reversing the argument, we expect to recover the VSI when the bouyancy
frequency becomes negligible compared to the cooling rate, or 
\begin{align}
  t_c \ll \frac{\Omega}{\gamma N_z^2}. 
\end{align} 
Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
for $N_z^2$ appropriate for vertically isothermal disks, we expect
\begin{align}
  \beta \ll
  \frac{1}{H^2_\mathrm{iso}\left(\gamma-1\right)\left(\frac{\p\ln\rho}{\p
          z}\right)^2} 
\end{align}
is required to enable the VSI, where we used $\Omega\sim\Omega_k$.  
This inequality is satisfied at the
midplane and as $|z|\to\infty$. It can be satisifed at all heights if
it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
Using the full expression for the density field for vertically isothermal
disks (Eq. \ref{vertiso_eqm}), we expect 
\begin{align}
  \beta \ll \frac{27}{4(\gamma-1)}\epsilon^2,
\end{align}
i.e. $\beta\ll O(\epsilon^2)$, is a sufficient condition for
buoyancy to be negligible.

@@

One effect of introducing finite thermal relaxation is buoyancy. % We
% show in the next section that buoyancy strongly stabilizes the
% VSI.
The $\zhat^2$ depedence in $\delta v_{z1}$ then makes sense because it
is most significant for large $\zhat$, i.e. away from the midplane
where the effect of buoyancy first appears as $t_c$ is increased.

@@

It is simplest to solve Eq. \ref{relax_disp} numerically. In
Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
($M=0$) VSI mode as a function of $k_xH_\mathrm{iso}$ for a range of
dimensionless thermal relaxation timescales $\beta$, in a disk with
$\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.  Similarly, 
in Fig. \ref{relax_bcrit} we solve the dispersion relation for the
critical $\beta$ such that the growth rate of the fundamental VSI mode
is zero.

Fig. \ref{relax_disp_fig}---\ref{relax_bcrit} shows that a growing fundamental
VSI mode exists both in the isothermal ($\beta\to0$) and adiabatic
($\beta\to\infty$) limits, although the latter growth rates are
smaller by at least a factor of two. Furthermore, the adiabatic VSI is 
is limited to a thin layer about the disk midplane so it may be
considered stable. Then Fig. \ref{relax_bcrit} shows that 
$\beta\lesssim 0.1$ is required for instability in this disk model. This critical
timescale is indepedent of $k_x$ for
$k_xH_\mathrm{iso}\gtrsim10$. Fig. \ref{relax_disp_fig} is also consistent with previous
discussions which show that thermal relaxation has no effect for
$k_x\to0$ or $k_x\to\infty$ as $\beta\to0$,

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_theory}
  \caption{Condition on the thermal relaxation timescale $\beta$ for the existence of unstable fundamental VSI
    modes in a vertically isothermal disk with $\epsilon=0.05$, $q=-1$
    and $\gamma=1.4$. For sufficiently small $\beta$ the isothermal VSI opertates,
    while for sufficiently large $\beta$ the adiabatic VSI
    operates. For intermediate values of $\beta$ the VSI growth rate
    is negative (i.e. stabilized). 
    \label{relax_bcrit}}  
\end{figure}

### 2014-12-29 linear_iso.tex
new fig for analytic discussion with therm relax

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the factor $d\Omega/dz$ in Eq. \ref{lin_vy}. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to the replacement
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

@@

For the vertically isothermal disk we have $|f(\hat{z})| =
|\hat{z}|\left(1+\epsilon^2\hat{z}^2\right)^{-3/2}$, which maximises
at $|\zhat|=1/\sqrt{2}\epsilon$. For vertical domains smaller 
than this height, the growth rate is limited by $|f|$ at the vertical
boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
\hat{z}$, in which case we can expect growth rates to increase linearly
with the vertical domain size.   %no faster than linear

>>>>

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to the replacement
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

@@

For the vertically isothermal disk we have $|f(\hat{z})| =
|\hat{z}|\left(1+\epsilon^2\hat{z}^2\right)^{-3/2}$, which maximises
at $|\zhat|=1/\sqrt{2}\epsilon$. For vertical domains smaller 
than this height, the growth rate is limited by $|f|$ at the vertical
boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
\hat{z}$, in which case we can expect growth rates to increase at most linearly
with respect to the vertical domain size.   %no faster than linear
